---
title: "Sensitivity Analysis for a Predator–Prey Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis for a Predator–Prey Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
```

## The Model

This vignette demonstrates how to define, compile, and integrate an ODE
system with **CppODE**, including automatic computation of parameter
sensitivities via forward-mode AD (FADBAD++).

As a simple example from theoretical ecology, we use the classic
**Lotka–Volterra** predator–prey model:

$$
\begin{aligned}
\dot{x} &= \alpha \, x - \beta \, x \, y \\
\dot{y} &= \delta \, x \, y - \gamma \, y
\end{aligned}
$$

where $x$ is the prey population and $y$ the predator population.

## Define and compile the model

```{r compile}
library(CppODE)
library(ggplot2)
library(tidyr)
library(dplyr)

eqns <- c(
  x = "alpha * x - beta * x * y",
  y = "delta * x * y - gamma * y"
)

# Compile with first-order sensitivities enabled
model <- CppODE(eqns, deriv = TRUE, modelname = "lotka_volterra")
```

## Solve the system

```{r solve}
parms <- c(x = 1.0, y = 0.5, alpha = 1.5, beta = 1.0, delta = 1.0, gamma = 3.0)
times <- seq(0, 15, length.out = 500)

res <- solveODE(model, times, parms)
diagnostics(res)
```

## State trajectories

```{r plot-states, fig.cap="Prey and predator population dynamics."}
df_states <- data.frame(
  time = res$time,
  x    = res$variable["x", ],
  y    = res$variable["y", ]
) |>
  pivot_longer(cols = c(x, y), names_to = "Species", values_to = "Population")

ggplot(df_states, aes(x = time, y = Population, colour = Species)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c(x = "#0072B2", y = "#D55E00"),
    labels = c(x = "Prey (x)", y = "Predator (y)")
  ) +
  labs(x = "Time", y = "Population", colour = NULL,
       title = "Lotka-Volterra Population Dynamics") +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## Phase portrait

```{r phase, fig.cap="Phase portrait in (x, y) space."}
df_phase <- data.frame(
  x = res$variable["x", ],
  y = res$variable["y", ]
)

ggplot(df_phase, aes(x = x, y = y)) +
  geom_path(colour = "#009E73", linewidth = 0.7) +
  geom_point(data = df_phase[1, ], colour = "black", size = 3) +
  annotate("text", x = df_phase$x[1] + 0.08, y = df_phase$y[1] + 0.05,
           label = "Start", hjust = 0, size = 3.5) +
  labs(x = "Prey (x)", y = "Predator (y)", title = "Phase Portrait") +
  theme_bw(base_size = 13) +
  coord_equal()
```

## Parameter sensitivities

The sensitivities $\partial x_i / \partial p_j$ are computed automatically
via forward-mode AD. `res$sens1` is a 3D array with dimensions
`[n_states, n_sens, n_times]`.

```{r plot-sensitivities, fig.height=7, fig.cap="First-order parameter sensitivities."}
sens_names <- dimnames(res$sens1)$sens
n_t <- length(res$time)

df_sens <- expand.grid(
  variable  = dimnames(res$sens1)$variable,
  parameter = sens_names,
  t_idx     = seq_len(n_t),
  stringsAsFactors = FALSE
)
df_sens$time  <- res$time[df_sens$t_idx]
df_sens$value <- as.vector(res$sens1)
df_sens$t_idx <- NULL

ggplot(df_sens, aes(x = time, y = value, colour = variable)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 2,
             labeller = labeller(parameter = setNames(
               paste0("\u2202\u00b7/\u2202", sens_names), sens_names
             ))) +
  scale_colour_manual(
    values = c(x = "#0072B2", y = "#D55E00"),
    labels = c(x = "\u2202x/\u2202p", y = "\u2202y/\u2202p")
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50",
             linewidth = 0.3) +
  labs(x = "Time", y = "Sensitivity", colour = "State",
       title = "First-Order Parameter Sensitivities") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        strip.background = element_rect(fill = "grey95"))
```

## Sensitivity to α in detail

How sensitive is the system to the prey growth rate $\alpha$?

```{r sens-alpha, fig.cap="Sensitivity of both populations to α."}
df_alpha <- data.frame(
  time      = res$time,
  dx_dalpha = res$sens1["x", "alpha", ],
  dy_dalpha = res$sens1["y", "alpha", ]
) |>
  pivot_longer(-time, names_to = "Sensitivity", values_to = "value")

ggplot(df_alpha, aes(x = time, y = value, colour = Sensitivity)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c(dx_dalpha = "#0072B2", dy_dalpha = "#D55E00"),
    labels = c(dx_dalpha = expression(partialdiff*x / partialdiff*alpha),
               dy_dalpha = expression(partialdiff*y / partialdiff*alpha))
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Time", y = "Sensitivity", colour = NULL,
       title = expression("Sensitivity w.r.t. " * alpha)) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## Summary

- **CppODE** generates and compiles C++ code for the ODE system at runtime.
- Sensitivities are computed automatically via forward-mode AD (FADBAD++) —
  no manual derivation of sensitivity equations required.
- Results are numerically exact (machine precision, no finite differences).

```{r session}
sessionInfo()
```
