---
title: "Cascade Signaling: Event-Driven Ligand Dosing with Sensitivities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cascade Signaling: Event-Driven Ligand Dosing with Sensitivities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
```

## The Model

This vignette walks through a realistic 15-state MAPK signaling cascade —
stiff, non-linear, with gene-regulatory feedback — and demonstrates two
features that go beyond the
[Predator–Prey vignette](CppODE_LotkaVolterra.html):

- **Time-triggered events**, used here to inject a ligand bolus into the
  ligand state at $t = t_{\text{dose}}$.
- **Sensitivity propagation across event boundaries** via the analytical
  **saltation matrix** that both backends apply automatically — so
  $\partial y / \partial p$ stays consistent through the dose jump.

The network is adapted from
`benchmarks/bench_cascade_model.R`, with the only difference that the
ligand `L` is promoted from a static parameter to a state variable so
that an event can act on it.

### States

The 15 states split into three functional layers:

| Layer | States |
|---|---|
| Receptor binding | `L`, `R`, `LR` |
| MAPK cascade | `KKK`, `KKKa`, `KK`, `KKa`, `K`, `Ka` |
| Gene regulation | `TF`, `TFa`, `mRNA_target`, `P_target`, `mRNA_inhib`, `P_inhib` |

Active phosphorylated kinases are suffixed with `a`. `P_inhib` closes a
negative feedback loop onto `KKKa` dephosphorylation.

## Define and compile the model

```{r define}
library(CppODE)
library(ggplot2)
library(tidyr)
library(dplyr)

rhs <- c(
  L           = "-kon * L * R + koff * LR",
  R           = "-kon * L * R + koff * LR + ksyn_R",
  LR          = " kon * L * R - koff * LR - kint * LR",
  KKK         = "-k1 * LR * KKK  / (Km1 + KKK)  + k2 * KKKa / (Km2 + KKKa) * (1 + ki * P_inhib)",
  KKKa        = " k1 * LR * KKK  / (Km1 + KKK)  - k2 * KKKa / (Km2 + KKKa) * (1 + ki * P_inhib)",
  KK          = "-k3 * KKKa * KK / (Km3 + KK)   + k4 * KKa  / (Km4 + KKa)",
  KKa         = " k3 * KKKa * KK / (Km3 + KK)   - k4 * KKa  / (Km4 + KKa)",
  K           = "-k5 * KKa * K   / (Km5 + K)    + k6 * Ka   / (Km6 + Ka)",
  Ka          = " k5 * KKa * K   / (Km5 + K)    - k6 * Ka   / (Km6 + Ka)",
  TF          = "-k7 * Ka * TF + k8 * TFa",
  TFa         = " k7 * Ka * TF - k8 * TFa",
  mRNA_target = "ktx   * TFa - kdeg_mRNA   * mRNA_target",
  P_target    = "ktl   * mRNA_target - kdeg_P   * P_target",
  mRNA_inhib  = "ktx_i * TFa - kdeg_mRNA_i * mRNA_inhib",
  P_inhib     = "ktl_i * mRNA_inhib  - kdeg_P_i * P_inhib"
)
```

### The event: add ligand at $t = 0$

We start with no ligand in the system (`L = 0`) and then add a bolus of
`L_dose` at time `t_dose` via a **fixed-time event**. Using a parameter
name for the time (`t_dose`) lets CppODE propagate sensitivities with
respect to the dose timing through the saltation matrix.

```{r event}
events <- data.frame(
  var    = "L",
  time   = "t_dose",
  value  = "L_dose",
  method = "add",
  root   = NA,
  stringsAsFactors = FALSE
)
```

### Compile

We compute sensitivities only for a handful of physiologically
interesting parameters — the dose magnitude `L_dose`, the receptor-binding
rate `kon`, the three phosphorylation rate constants `k1`/`k3`/`k5`, the
transcription-factor activation rate `k7`, and the feedback strength
`ki`. Every other parameter and initial condition is declared `fixed`,
which turns them into plain scalars at the C++ level and skips their AD
columns entirely.

```{r compile}
all_states <- names(rhs)
sens_pars  <- c("L_dose", "kon", "k1", "k3", "k5", "k7", "ki")
all_pars   <- c("t_dose", "L_dose", "kon", "koff", "kint", "ksyn_R",
                "k1", "Km1", "k2", "Km2", "k3", "Km3", "k4", "Km4",
                "k5", "Km5", "k6", "Km6", "ki", "k7", "k8",
                "ktx", "kdeg_mRNA", "ktl", "kdeg_P",
                "ktx_i", "kdeg_mRNA_i", "ktl_i", "kdeg_P_i")
fixed <- c(all_states, setdiff(all_pars, sens_pars))

model <- CppODE(
  rhs,
  events    = events,
  fixed     = fixed,
  deriv     = TRUE,
  method    = "bdf",
  useNDF    = TRUE,
  modelname = "cascade_signaling",
  outdir    = tempdir()
)
```

The default multistep is NDF-on-BDF — a good stiff choice for this
network. For comparison you can swap `method = "rb4"` (Rosenbrock-4).

## Solve

```{r solve}
parms <- c(
  # Initial conditions
  L = 0, R = 100, LR = 0,
  KKK = 100, KKKa = 0, KK = 100, KKa = 0, K = 100, Ka = 0,
  TF = 100, TFa = 0,
  mRNA_target = 0, P_target = 0, mRNA_inhib = 0, P_inhib = 0,
  # Event parameters
  t_dose = 0.1, L_dose = 10,
  # Kinetics
  kon = 0.1, koff = 0.05, kint = 0.01, ksyn_R = 0.5,
  k1 = 0.5, Km1 = 10, k2 = 0.2, Km2 = 10,
  k3 = 0.3, Km3 = 10, k4 = 0.15, Km4 = 10,
  k5 = 0.4, Km5 = 10, k6 = 0.1, Km6 = 10, ki = 0.5,
  k7 = 0.2, k8 = 0.05,
  ktx = 1.0, kdeg_mRNA = 0.1, ktl = 0.5, kdeg_P = 0.05,
  ktx_i = 0.3, kdeg_mRNA_i = 0.2, ktl_i = 0.4, kdeg_P_i = 0.1
)

times <- seq(0, 100, length.out = 500)
res   <- solveODE(model, times, parms, abstol = 1e-9, reltol = 1e-7)
diagnostics(res)
```

`t_dose = 0.1` places the event just inside the integration interval —
the jump is visible at the very start of the plots below.

## Cascade dynamics

```{r plot-cascade, fig.height=6, fig.cap="State trajectories after the ligand bolus. Dashed vertical line marks t = t_dose."}
df <- cbind(time = res$time, as.data.frame(res$variable)) |>
  pivot_longer(-time, names_to = "state", values_to = "value")

layer <- c(
  L = "Receptor",  R = "Receptor",  LR = "Receptor",
  KKK = "Kinase",  KKKa = "Kinase", KK = "Kinase", KKa = "Kinase",
  K = "Kinase",    Ka = "Kinase",
  TF = "Gene",     TFa = "Gene",
  mRNA_target = "Gene", P_target = "Gene",
  mRNA_inhib  = "Gene", P_inhib  = "Gene"
)
df$layer <- factor(layer[df$state],
                   levels = c("Receptor", "Kinase", "Gene"))

ggplot(df, aes(x = time, y = value, colour = state)) +
  geom_vline(xintercept = parms["t_dose"], linetype = "dashed",
             colour = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ layer, scales = "free_y", ncol = 1) +
  labs(x = "Time", y = "Concentration", colour = NULL,
       title = "Cascade response to ligand bolus") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right",
        strip.background = element_rect(fill = "grey95"))
```

The ligand is consumed by receptor binding, the cascade transduces the
signal upward through `KKKa → KKa → Ka`, and the gene-regulatory layer
lags behind with both a target product (`P_target`) and a feedback
inhibitor (`P_inhib`) that eventually dampens the active pool.

## Parameter sensitivities

`res$sens1` is a 3D array with dimensions `[n_times, n_states, n_sens]`.
Because we declared most parameters `fixed`, the sensitivity dimension
collapses from the 43 total symbols down to the seven we actually care
about.

```{r sens-summary}
dim(res$sens1)
dimnames(res$sens1)$sens
```

### Sensitivity of active MAPK (`Ka`) to kinetic parameters

```{r plot-sens, fig.cap="How much does each kinetic parameter drive the active MAPK (Ka) response?"}
df_sens <- data.frame(time = res$time, res$sens1[, "Ka", ]) |>
  pivot_longer(-time, names_to = "parameter", values_to = "value")

ggplot(df_sens, aes(x = time, y = value, colour = parameter)) +
  geom_vline(xintercept = parms["t_dose"], linetype = "dashed",
             colour = "grey60", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey70") +
  geom_line(linewidth = 0.7) +
  labs(x = "Time", y = expression(partial*Ka / partial*p),
       colour = "p",
       title = "First-order sensitivity of Ka") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")
```

Notice the jump in $\partial Ka / \partial L_{\text{dose}}$ at
$t = t_{\text{dose}}$: the event rewrites `L`, and the saltation matrix
propagates that perturbation into every downstream state's sensitivity
coordinates — **no finite differences**, no manual event-aware
chain rule.

### Comparison across output layers

```{r plot-sens-outputs, fig.height=6, fig.cap="Sensitivity of one representative state per layer to the ligand dose."}
outputs <- c(Receptor = "LR", Kinase = "Ka", Gene = "P_target")

df_out <- do.call(rbind, lapply(names(outputs), function(lyr) {
  st <- outputs[[lyr]]
  data.frame(time  = res$time,
             layer = lyr,
             state = st,
             value = res$sens1[, st, "L_dose"])
}))
df_out$layer <- factor(df_out$layer, levels = names(outputs))

ggplot(df_out, aes(x = time, y = value)) +
  geom_vline(xintercept = parms["t_dose"], linetype = "dashed",
             colour = "grey60", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey70") +
  geom_line(linewidth = 0.7, colour = "#0072B2") +
  facet_wrap(~ layer + state, scales = "free_y", ncol = 1) +
  labs(x = "Time", y = expression(partial*y / partial*L[dose]),
       title = "Dose sensitivity propagates down the cascade") +
  theme_bw(base_size = 12) +
  theme(strip.background = element_rect(fill = "grey95"))
```

The dose sensitivity rises immediately at the receptor layer, peaks a
few time units later at the kinase layer, and builds slowly at the gene
layer — the characteristic temporal hierarchy of signaling cascades,
recovered directly from first-order AD.

## Cross-check with the CVODE backend

Both `CppODE()` and `CVODE()` share the same front-end. Swapping the
constructor gives a second, independently compiled solver with sparse
KLU linear algebra — and the two agree to integration tolerance.

```{r cvode-available, include=FALSE}
have_cvode <- isTRUE(CppODE:::cvodeConfig$available)
```

```{r cvode, eval=have_cvode}
model_cv <- CVODE(
  rhs,
  events    = events,
  fixed     = fixed,
  deriv     = TRUE,
  method    = "bdf",
  sparse    = TRUE,
  modelname = "cascade_signaling_cvode",
  outdir    = tempdir()
)

res_cv <- solveODE(model_cv, times, parms, abstol = 1e-9, reltol = 1e-7)

# Align on the shared user time grid (CppODE inserts the event time as an
# extra row; CVODE does not — pick the times both reported).
keep_cpp <- res$time    %in% times
keep_cv  <- res_cv$time %in% times

max_state_diff <- max(abs(res$variable[keep_cpp, ] - res_cv$variable[keep_cv, ]))
max_sens_diff  <- max(abs(res$sens1[keep_cpp, , ]  - res_cv$sens1[keep_cv, , ]))

cat(sprintf("  L-infinity difference (states)       : %.3e\n", max_state_diff))
cat(sprintf("  L-infinity difference (sensitivities): %.3e\n", max_sens_diff))
```

```{r cvode-skip, eval=!have_cvode, echo=FALSE, results='asis'}
cat(
  "> **Note:** the CVODE cross-check was skipped because the CVODE ",
  "backend is disabled in this installation — SUNDIALS (>= 6.0) was ",
  "not found on the build host. Install `libsundials-dev` ",
  "(Debian/Ubuntu), `sundials-devel` (Fedora), `brew install sundials` ",
  "(macOS), or on Windows run ",
  "`C:/rtools<ver>/usr/bin/pacman.exe -Sy --noconfirm mingw-w64-ucrt-x86_64-sundials` ",
  "(substitute your installed Rtools version for `<ver>`). ",
  "Then re-install CppODE to enable it.\n",
  sep = ""
)
```

## Summary

- A single **event data frame** (`var`, `time`, `value`, `method`,
  `root`) is all that is needed to inject a dose — the backend merges the
  event time into the integration grid, applies the action, and resets
  the stepper.
- **Sensitivities and events interact cleanly** thanks to the analytical
  saltation-matrix correction. First-order AD costs one extra state per
  sensitivity parameter — the exact number you get back, no more.
- **CppODE and CVODE are interchangeable.** Use the CppODE multistepper
  for pure R-side control and deep integration with the AD chain rule;
  use CVODE when you want SUNDIALS' battle-tested BDF or very large
  sparse problems through KLU.

For a simpler starting point without events see
`vignette("CppODE_LotkaVolterra", package = "CppODE")`.

```{r session}
sessionInfo()
```
